Still Missing a Cool Title

### 2006-05-31 talk/dls2006/paper.tex
Fix figure 1.

\section{System programming with Python}
\label{systemprog}

@@

This means that in addition to transforming the existing graphs, each
transformation step also needs to insert new functions into the forest.
A key feature of our approach is that we can write such "system-level"
code -- relevant only to a particular transformation -- in plain Python
as well:

\begin{verbatim}
.. topic:: Figure 1 - a helper to implement \texttt{list.append()}

  ::

    def ll_append(lst, newitem):
        # Append an item to the end of the vector.
        index = lst.length         # get the 'length' field
        ll_resize(lst, index+1)    # call a helper not shown here
        itemsarray = lst.items     # get the 'items' field
        itemsarray[index] = item   # this behaves like a C array
\end{verbatim}

The idea is to feed these new Python functions into the front-end, using
this time the transformation's target (lower-level) type system during
the type inference.  In other words, we can write plain Python code that
manipulates objects that conform to the lower-level type system, and
have these functions automatically transformed into appropriately typed
graphs.

For example, \texttt{ll\textunderscore{}append()} in figure 1 is a Python function
that manipulates objects that behave like C structures and arrays.
This function is inserted by the LLTyper, as a helper to implement the
\texttt{list.append()} calls found in its RPython-level input graphs.
By going through the front-end reconfigured to use C-level types, the
above function becomes a graph with such C-level types\footnote{The
low-level type system specifies that the function should be
specialized by the C-level type of its input arguments, so it actually
turns into one graph per list type -- list of integers, list of
pointers, etc.  This behavior gives the programmer a feeling
comparable to C++ templates, without the declarations.}, which is then
indistinguishable from the other graphs of the forest produced by the
LLTyper.

@@

{\bf pypy-c-stackless-gcframework.}
    This variant combines the "gc transformation" step with the
    "stackless transformation" step.  The overhead introduced by the
    stackless feature is theoretically balanced with the removal of the
    overhead of pushing and popping roots explicitly on an alternate
    stack: indeed, in this variant it is possible to ask the functions
    in the current C call chain to save their local state and return.
    This has the side-effect of moving all roots to the heap, where the
    GC can find them.  (We hypothesize that the large slowdown is caused
    by the extreme size of the executable in this case - 21MB, compared to
    6MB for the basic pypy-c.  Making it smaller is work in progress.)

@@

Applying the expressiveness - or at least the syntax - of very
high-level and dynamically typed languages to their implementation has
been investigated many times.

@@

Jikes RVM [DBLP:journals/ibmsj/AlpernABBCCCFGHHLLMNRSSSSSSW00] is a
Java VM and Just-In-Time compiler written in Java.
Bootstrapping happens by self-applying the compiler on a host VM, and
dumping a snapshot from memory of the resulting native code.
This approach directly enables high performance, at the price of
portability - as usual with pure native code emitting
approaches. Modularity of features, when possible, is achieved with
normal software modularity. The indirection costs are taken care of by
the compiler performing inlining (which is sometimes even explicitly
requested).  In particular this modular approach is used
for implementing a range of choices for GC support [Jikes-GC].  This was
the inspiration for PyPy's own GC framework, although much more tuning
and work went into Jikes RVM.  PyPy's own GC framework also exploits
inlining of helpers and barriers to recover performance.

@@

A desirable property of our approach is to allow a given language and VM
to be specified only once, in the form of an interpreter.  Moreover, the
interpreter can be kept simple (and thus keep its role as a
specification): not only is it written in a high-level language, but it
is not overloaded with low-level design choices and implementation
details.  This makes language evolution and experimentation easier.
More generally, this property is important because many interpreters for
very difference languages can be written: the simpler these interpreters
can be kept, the more we win from our investment in writing the
tool-chain itself - a one-time effort.

>>>>

\section{System programming in Python}
\label{systemprog}

@@

This means that in addition to transforming the existing graphs, each
transformation step also needs to insert new functions into the forest.
A key feature of our approach is that we can write such "system-level"
code -- relevant only to a particular transformation -- in plain Python
as well.  The idea is to feed these new Python functions into the
front-end, using this time the transformation's target (lower-level)
type system during the type inference.  In other words, we can write
plain Python code that manipulates objects that conform to the
lower-level type system, and have these functions automatically
transformed into appropriately typed graphs.

For example, \texttt{ll\textunderscore{}append()} in figure \ref{llappend}
is a Python function
that manipulates objects that behave like C structures and arrays.
This function is inserted by the LLTyper, as a helper to implement the
\texttt{list.append()} calls found in its RPython-level input graphs.
By going through the front-end reconfigured to use C-level types, the
above function becomes a graph with such C-level types\footnote{The
low-level type system specifies that the function should be
specialized by the C-level type of its input arguments, so it actually
turns into one graph per list type -- list of integers, list of
pointers, etc.  This behavior gives the programmer a feeling
comparable to C++ templates, without the declarations.}, which is then
indistinguishable from the other graphs of the forest produced by the
LLTyper.

\begin{figure}
\begin{verbatim}
def ll_append(lst, newitem):
  # Append an item to the end of the vector.
  index = lst.length       # get the 'length' field
  ll_resize(lst, index+1)  # call another helper
  itemsarray = lst.items   # get the 'items' field
  itemsarray[index] = item # behaves like a C array
\end{verbatim}
\caption{A helper to implement \texttt{list.append()}.}
\label{llappend}
\end{figure}

@@

{\bf pypy-c-stackless-gcframework.}
    This variant combines the "gc transformation" step with the
    "stackless transformation" step.  The overhead introduced by the
    stackless feature is theoretically balanced with the removal of the
    overhead of pushing and popping roots explicitly on an alternate
    stack: indeed, in this variant it is possible to ask the functions
    in the current C call chain to save their local state and return.
    This has the side-effect of moving all roots to the heap, where the
    GC can find them.  (We hypothesize that the large slowdown is caused
    by the extreme size of the executable in this case -- 21MB, compared to
    6MB for the basic pypy-c.  Making it smaller is work in progress.)

@@

Applying the expressiveness -- or at least the syntax -- of very
high-level and dynamically typed languages to their implementation has
been investigated many times.

@@

Jikes RVM [DBLP:journals/ibmsj/AlpernABBCCCFGHHLLMNRSSSSSSW00] is a
Java VM and Just-In-Time compiler written in Java.
Bootstrapping happens by self-applying the compiler on a host VM, and
dumping a snapshot from memory of the resulting native code.
This approach directly enables high performance, at the price of
portability -- as usual with pure native code emitting
approaches. Modularity of features, when possible, is achieved with
normal software modularity. The indirection costs are taken care of by
the compiler performing inlining (which is sometimes even explicitly
requested).  In particular this modular approach is used
for implementing a range of choices for GC support [Jikes-GC].  This was
the inspiration for PyPy's own GC framework, although much more tuning
and work went into Jikes RVM.  PyPy's own GC framework also exploits
inlining of helpers and barriers to recover performance.

@@

A desirable property of our approach is to allow a given language and VM
to be specified only once, in the form of an interpreter.  Moreover, the
interpreter can be kept simple (and thus keep its role as a
specification): not only is it written in a high-level language, but it
is not overloaded with low-level design choices and implementation
details.  This makes language evolution and experimentation easier.
More generally, this property is important because many interpreters for
very difference languages can be written: the simpler these interpreters
can be kept, the more we win from our investment in writing the
tool-chain itself -- a one-time effort.
